Making ice sheet models \\ scale properly

### 2023-02-10 talk/slides.tex
the talk I will give

\bigskip
\begin{minipage}[t][60cm][t]{\textwidth}
\begin{center}
\includegraphics<1>[height=0.69\textheight]{images/ant-pittard2021.png}
\only<1>{\par {\scriptsize Antarctic ice sheet}} % (Pittard et al 2021)
\only<2>{\vspace{17mm}}
\includegraphics<2>[width=\textwidth]{images/ant-schoofhewitt2013.png}
\only<2>{\vspace{13mm}}
\only<2>{\par {\scriptsize note vertical exaggeration and rough bed (Schoof \& Hewitt 2013)}}
\includegraphics<3>[height=0.69\textheight]{images/alps-seguinot2018.png}
\only<3>{\par {\scriptsize modeled Alpine ice sheet near last glacial maximum (Seguinot et al 2018)}}
\includegraphics<4>[height=0.69\textheight]{images/british-clark2022.png}
\only<4>{\par {\scriptsize modeled British-Irish ice sheet near last glacial maximum (Clark et al 2022)}}
\includegraphics<5>[height=0.69\textheight]{images/not-sea-ice.png}
\only<5>{\par {\scriptsize \emph{not} sea ice!}}
\end{center}
\end{minipage}
\end{frame}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item glacier ice is modeled as a \emph{slow, incompressible, viscous, non-Newtonian fluid}
    \begin{itemize}
    \item[$\circ$] more on that soon
    \end{itemize}
\item glaciers lie on \emph{topography}, and sometimes they float  (\emph{ice shelf})
\item glacier geometry and velocity evolve \emph{in contact with climate}:
    \begin{itemize}
    \item[$\circ$] snowfall
    \item[$\circ$] surface melt
    \item[$\circ$] subglacial melt
    \item[$\circ$] sub-shelf melt (when floating)
    \item[$\circ$] calving (into ocean)
    \end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item for simplicity/clarity of the upcoming model performance analysis, I will ignore much of glacier physics
\item \alert{ignoring}:
    \begin{itemize}
    \item[$\circ$] floating ice
    \item[$\circ$] subglacial hydrology
    \item[$\circ$] ice temperature
    \item[$\circ$] fracture processes (calving, crevasses)
    \item[$\circ$] solid earth deformation
    \end{itemize}

\medskip
\item<2> {\footnotesize see UAF's \href{https://pism.io/}{Parallel Ice Sheet Model (\texttt{pism.io})}, for example, as a model which includes these processes}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

@@

\begin{columns}
\begin{column}{0.55\textwidth}
\begin{itemize}
\item data $a(t,x,y)$, $b(x,y)$  is defined on a \alert{fixed domain}:
	$$t \in [0,T] \quad \text{and} \quad (x,y) \in \Omega \subset \RR^2$$
\item<2-> solution surface elevation $s(t,x,y)$ is defined on $[0,T]\times \Omega$
    \begin{itemize}
    \item[$\circ$] also a fixed domain,
    \item[$\circ$] but $s=b$ where there is no ice
    \end{itemize}
\item<3> one must \emph{solve} for the time-dependent \alert{icy domain} $\Lambda(t) \subset \RR^3$, on which (solution) velocity $\bu(t,x,y,z)$ is defined:
    $$\Lambda(t) = \left\{(x,y,z)\,:\,b(x,y) < z < s(t,x,y)\right\}$$

@@

        \begin{itemize}
        \item[$\vartriangleright$] $\Gamma_s(t), \Gamma_b(t) \subset \partial \Lambda(t)$ denote the surface and base of the ice
        \end{itemize}
    \end{itemize}

@@

\begin{itemize}
\item \only<1>{\alert{nonlinear complementarity problem} (NCP)} \only<2>{nonlinear complementarity problem (NCP) coupled to \alert{Stokes}}:
\begin{align*}
s - b &\ge 0 && \text{on $\Omega \subset \RR^2$} \\
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &\ge 0 && \text{''} \\
(s - b) \left(\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a\right) &= 0 && \text{''} \\
\onslide<2->{- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda(t) \subset \RR^3$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b(t)$} \\
\bu|_b \cdot \bn_b &= 0 && \text{''} \\
\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s(t)$}}
\end{align*}

    \begin{itemize}
    \item[$\circ$] $\bu|_s=0$ where no ice
    \item<2->[$\circ$] viscosity by Glen law: \, $2\nu_\eps(D\bu) = \Gamma \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item for this slide, forget complementarity and boundary conditions
\item result: SKE coupled to Stokes
\begin{align*}
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &= 0 \\
- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero \\
\nabla \cdot \bu &= 0
\end{align*}

@@

\begin{itemize}
\item to the best of my knowledge, \emph{no} current research groups are studying well-posedness or regularity for this basic model
    \begin{itemize}
    \item[$\circ$] when pressed, most researchers would agree NCP-coupled-to-Stokes is the \emph{intended} model?
    \item[$\circ$] well-posedness of the lubrication approximation of the model has been considered; existence proved in (Jouvet \& Bueler 2012)
    \end{itemize}

@@

\medskip
\item NCP-coupled-to-Stokes is \emph{not yet} in common use for high-resolution, long-duration ice sheet simulations
    \begin{itemize}
    \item[$\circ$] \alert{can we make it fast enough to use?}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item a non-shallow model solves a Stokes problem at each step:
\begin{align*}
- \nabla \cdot \left(2 \nu_\eps(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda \subset \RR^3$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b$} \\
\bu|_b \cdot \bn_b &= 0 && \text{''} \\
\left(2 \nu_\eps(D\bu) D\bu - pI\right) \bn &= \bzero && \text{on $\Gamma_s$}
\end{align*}
\item this is the \alert{stress balance} (conservation of momentum) problem which determines velocity $\bu$ and pressure $p$
\item how fast is the numerical solution process?
    \begin{itemize}
    \item[$\circ$] how do solution algorithms \alert{scale} with increasing spatial resolution?
    \end{itemize}
\end{itemize}
\end{frame}

>>>>

\bigskip
\begin{minipage}[t][60cm][t]{\textwidth}
\begin{center}
\includegraphics<1>[height=0.69\textheight]{images/ant-pittard2021.png}
\only<1>{\par {\scriptsize Antarctic ice sheet}} % (Pittard et al 2021)
\only<2>{\vspace{17mm}}
\includegraphics<2>[width=\textwidth]{images/ant-schoofhewitt2013.png}
\only<2>{\bigskip \hfill {\tiny (Schoof \& Hewitt 2013)}}
\only<2>{\vspace{4mm}}
\only<2>{\begin{center} {\footnotesize note vertical exaggeration, smooth surface, and rough bed} \end{center} }
\includegraphics<3>[height=0.69\textheight]{images/alps-seguinot2018.png}
\only<3>{\par {\scriptsize modeled Alpine ice sheet near last glacial maximum} \hfill {\tiny (Seguinot et al 2018)}}
\includegraphics<4>[height=0.69\textheight]{images/british-clark2022.png}
\only<4>{\par {\scriptsize modeled British-Irish ice sheet near last glacial maximum} \hfill {\tiny  (Clark et al 2022)}}
\includegraphics<5>[height=0.69\textheight]{images/not-sea-ice.png}
\only<5>{\par {\footnotesize an ice sheet is \emph{not} sea ice!}}
\end{center}
\end{minipage}
\end{frame}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item glacier ice is modeled as a \emph{very viscous, incompressible, non-Newtonian fluid}
    \begin{itemize}
    \item[$\circ$] more on that soon
    \end{itemize}
\item glaciers lie on \emph{topography}
    \begin{itemize}
    \item[$\circ$] sometimes they float  (\emph{ice shelf})
    \end{itemize}
\item glacier geometry and velocity evolve \emph{in contact with climate}:
    \begin{itemize}
    \item[$\circ$] snowfall
    \item[$\circ$] surface melt
    \item[$\circ$] subglacial melt
    \item[$\circ$] sub-shelf melt (when floating)
    \item[$\circ$] calving (into ocean)
    \end{itemize}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

@@

\begin{columns}
\begin{column}{0.6\textwidth}
\begin{itemize}
\item for simplicity/clarity of the upcoming modeling, I will ignore much of glacier physics
\item \alert{ignoring}:
    \begin{itemize}
    \item[$\circ$] floating ice
    \item[$\circ$] subglacial hydrology
    \item[$\circ$] ice temperature
    \item[$\circ$] fracture processes (calving, crevasses)
    \item[$\circ$] solid earth deformation
    \end{itemize}

\medskip
\item {\footnotesize see UAF's \href{https://pism.io/}{Parallel Ice Sheet Model (\texttt{pism.io})}, for example, as a model which includes these processes}
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\hfill \includegraphics[width=0.9\textwidth]{images/map-glacier-ice.png}
\end{column}
\end{columns}
\end{frame}

@@

\begin{columns}
\begin{column}{0.55\textwidth}
\begin{itemize}
\item data $a(t,x,y)$, $b(x,y)$ are defined on a \alert{fixed domain}:
	$$t \in [0,T] \quad \text{and} \quad (x,y) \in \Omega \subset \RR^2$$
\item<2-> solution surface elevation $s(t,x,y)$ is defined on $[0,T]\times \Omega$
    \begin{itemize}
    \item[$\circ$] also a fixed domain,
    \item[$\circ$] but $s=b$ where there is no ice
    \end{itemize}
\item<3> $s(t,x,y)$ determines the time-dependent \alert{icy domain} $\Lambda(t) \subset \RR^3$, on which the solution velocity $\bu(t,x,y,z)$ is defined:
    $$\Lambda(t) = \left\{(x,y,z)\,:\,b(x,y) < z < s(t,x,y)\right\}$$

@@

        \begin{itemize}
        \item[$\vartriangleright$] $\Gamma_s(t), \Gamma_b(t) \subset \partial \Lambda(t)$ denote the surface and base of the ice
        \item[$\vartriangleright$] $\bn_s = \left<-\grad s,1\right>$ is upward surface normal
        \end{itemize}
    \end{itemize}

@@

\begin{itemize}
\item \only<1>{\alert{nonlinear complementarity problem} (NCP)} \only<2>{nonlinear complementarity problem (NCP) coupled to \alert{Stokes}}:
\begin{align*}
s - b &\ge 0 && \text{on $\Omega \subset \RR^2$} \\
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &\ge 0 && \text{''} \\
(s - b) \left(\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a\right) &= 0 && \text{''} \\
\onslide<2->{- \nabla \cdot \left(2 \nu(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda(t) \subset \RR^3$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b(t)$} \\
\bu|_b \cdot \bn_b &= 0 && \text{''} \\
\left(2 \nu(D\bu) D\bu - pI\right) \bn_s &= \bzero && \text{on $\Gamma_s(t)$}}
\end{align*}

    \begin{itemize}
    \item[$\circ$] $\bu|_s=0$ where no ice
    \item<2->[$\circ$] viscosity by Glen law: \, $2\nu(D\bu) = \Gamma |D\bu|^{\pp-2}$
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item for this slide, forget complementarity and boundary conditions
\item result: SKE coupled to Stokes
\begin{align*}
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &= 0 \\
- \nabla \cdot \left(2 \nu(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero \\
\nabla \cdot \bu &= 0
\end{align*}

@@

\begin{itemize}
\item to the best of my knowledge, \emph{no} current research groups are studying well-posedness or regularity for this basic model
    \begin{itemize}
    \item[$\circ$] when pressed, most researchers would agree NCP-coupled-to-Stokes is the \emph{intended} model
    \item[$\circ$] well-posedness of the lubrication approximation of the model has been considered; existence proved in (Jouvet \& Bueler 2012)
    \end{itemize}

@@

\medskip
\item NCP-coupled-to-Stokes is \emph{not yet} in common use for high-resolution, long-duration ice sheet simulations
    \begin{itemize}
    \item[$\circ$] because it is too slow
    \item[$\circ$] \alert{can we make it fast enough to use?}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item a non-shallow model solves a Stokes problem at each step:
\begin{align*}
- \nabla \cdot \left(2 \nu(D\bu)\, D\bu\right) + \nabla p - \rhoi \mathbf{g} &= \bzero && \text{in $\Lambda \subset \RR^3$} \\
\nabla \cdot \bu &= 0 && \text{''} \\
\btau_b - \bbf(\bu|_b) &= \bzero && \text{on $\Gamma_b$} \\
\bu|_b \cdot \bn_b &= 0 && \text{''} \\
\left(2 \nu(D\bu) D\bu - pI\right) \bn_s &= \bzero && \text{on $\Gamma_s$}
\end{align*}
\item this is the \alert{stress balance} (conservation of momentum) problem which determines velocity $\bu$ and pressure $p$
\item how fast is the numerical solution process?
    \begin{itemize}
    \item[$\circ$] how do solution algorithms \alert{scale} with increasing spatial resolution?
    \end{itemize}
\end{itemize}
\end{frame}
